# BeginneRs {#beginners}

> Working with data is a core skill for researchers, and increasingly important
> in all professional careers.

> There are many tools for manpulating, visualising and analysing data. R is
> powerful, open-source tool which has gained wide acceptance among researchers
> and in industry, and expertise with R is highly sought after by employers.

> In this course we'll learn R as we go, building a little at a time. This and
> the next session covers the basics (things you would learn as part of an
> undergraduate statistics course at Plymouth). If some of this is familiar,
> there will be extension exercises to enhance your knowledge.

## Session 1

Outline:

-   R accounts handout and login TODO PLAN
-   TODO ADD ANDY's initial worksheets

```{r}
library(webex)
library(tidyverse)
```

Example of a question. `r 2+2`  
What is your favourite letter? `r mcq(c("A", "B", "C", answer="D"))`

```{r, results="asis"}
mcq(c("A", "B", "C", answer="D")) %>% cat
```

Hints

`r hide("Show hint")`

To mimic the plot you will need to use a `geom_boxplot()` and also
`facet_grid(~sex)`.

`r unhide()`

## Session 2

TBC

<!-- add from Andy -->
