---
title: 'Data wrangling: Extension Exercises'
author: 'Ben Whalley'
date: "September 2020"
bibliography: [references.bib]
biblio-style: apa6
link-citations: yes
output: 
  webex::html_clean:
    includes:
      after_body: footer.html
---


```{r, echo=F, include=F}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE, comment=">", message=FALSE)
library(tidyverse)
library(webex)
library(pander)
theme_set(theme_minimal())
```



# More on Long vs wide data


## Further reading

:::{.exercise}

**#1**

To gain more understanding of the way the format of our data is important to the way we analyse it, read this article on wide/long data formats and the tradeoffs between the two: <https://www.theanalysisfactor.com/wide-and-long-data/>

:::




## Separating 'untidy' variables into tidy, long-form data:

The code below generates simulated data for 100 individuals at three time points. The format is
similar to the way you might record experimental data in a spreadsheet.

```{r}
N <- 20
repeatmeasuresdata <- tibble(person = 1:N,
  time_1 = rnorm(N),
  time_2 = rnorm(N, 1),
  time_3 = rnorm(N, 3))

repeatmeasuresdata %>% head
```

`repeatmeasuresdata` is in **wide** format. Each row contains data for one participant, and each
participant has three observations.

We can melt or reshape the data into longer format like so:

```{r}
repeatmeasuresdata %>%
  pivot_longer(starts_with("time")) %>%
  arrange(person, name) %>%
  head(8)
```

The problem we have now is that `name` contains text which describes at which time the
observation was made. We probably want a number for each timepoint, so we can make a plot with time
on the x axis.

The `separate` command separates a single character column (`name`) into multiple columns.
Rather than have a column with labels of the form 'time_1', it can create two columns, with labels
'time' and '1' in each.

```{r}
longrepeatmeasuresdata <- repeatmeasuresdata %>%
  pivot_longer(starts_with("time")) %>%
  separate(name, into = c("variable", "time"))

longrepeatmeasuresdata %>% head
```


Now the data are in long format, we can plot the points over time:

```{r}
longrepeatmeasuresdata %>%
  sample_n(30) %>%
  ggplot(aes(x=time, y=value)) +
  geom_point()
```



:::{.exercise}

**#2**

This file contains sample contact and address data for 100 people: https://letterhub.com/wp-content/uploads/2018/03/100-contacts.csv


- Read the data into R
- Use the `separate` function to make a new variable which contains the *domain name* of these contacts' email address (e.g. yahoo.com, hotmail.com)

Use the `distinct` and/or `count` functions on the new variable you create containing the domain name. Look them up in the help file if you don't know which to use to answer these questions:

- How many different domains do the contacts have addresses with?
- How many people had a gmail account?
- Which domains had more than 10 users?


`r hide("Show workings")`


```{r}
contacts <- read_csv('https://letterhub.com/wp-content/uploads/2018/03/100-contacts.csv') %>% 
  separate(email, into=c("user", "domain"), sep ="@") 


# how many different domains
contacts %>% 
  distinct(domain) %>% 
  count()

# how many people use gmail
contacts %>% 
  count(domain) %>% 
  filter(domain=="gmail.com")

# which domains had more than 10 users?
contacts %>% 
  count(domain) %>% 
  filter(n > 10) 
```


`r unhide()`





:::






# More `pivot_longer` and plotting


:::{.exercise}

**#3**

Use the `mtcars` data and facetting to produce these plots:

```{r, echo=F, message=F, warning=F}
mtcars %>%
  select(mpg, wt, disp) %>%
  pivot_longer(everything()) %>% 
  mutate(variable=Hmisc::capitalize(as.character(name))) %>%
  ggplot(aes(value, y=..scaled..)) +
  geom_density() +
  facet_wrap(~variable, scales="free_x") +
  labs(x="", y="Scaled density")
```

```{r, echo=F}
mtcars %>%
  select(mpg, wt, disp) %>%
  pivot_longer(everything()) %>% 
  mutate(variable=Hmisc::capitalize(as.character(name))) %>%
  ggplot(aes(value)) +
  geom_histogram(bins=7) +
  facet_wrap(~variable, scales="free_x") +
  labs(x="", y="Scaled density")
```

Some tips:

-   Add `scales="free_x"` to the `facet_wrap` function. So, it will read:
    `facet_wrap(~VARNAME, scales="free_x")`

-   For the histogram, replace `geom_density()` with `geom_histogram()`

-   Adjust the number of vertical bars in the histogram by adding `bins=7` (or some other number) to
    `geom_histogram()`

:::






# More practice

For more practice using these techniques for common problems in psychological research you could work through Andy Will's 'better tables' worksheet here: 

- <https://ajwills72.github.io/rminr/better-tables.html>
- [Preprocessing data from experiments](https://ajwills72.github.io/rminr/preproc.html) and [part 2 of this](https://ajwills72.github.io/rminr/more-on-preproc.html)



------------


All content on this site distributed under a [Creative Commons](https://creativecommons.org/)
licence. CC-BY-SA 4.0.
