---
title: 'Multiple regression'
author: 'Ben Whalley'
date: "September 2020"
bibliography: [references.bib]
biblio-style: apa6
link-citations: yes
output:
  webex::html_clean:
    includes:
      after_body: footer.html
---



```{r include=F,echo=F}
library(tidyverse)
library(webex)
library(cowplot)
library(DiagrammeR)
library(pander)
library(broom)
library(plotly)
source('grvizpng.R')
theme_set(theme_minimal())
knitr::opts_chunk$set(cache=T, message=F, warning=F)
```

```{r include=F,echo=F, cache=T}
studyhabits <- read_csv('https://benwhalley.github.io/rmip/data/studyhabitsandgrades.csv')
```




NOTE: if you are not sure why the following exercises are useful or relevant to the multiple regression session then bear with me to the end; it will become clearer!



# Sequences and designs for sampling and allocation

One trick you may find useful later in the course is making **sequences** of numbers.

There are a few ways to do this, but the simplest is to write: `1:10`. That is, the number to start
from (`1`), a colon (`:`), and then the number to end with (`10`).

Copy and paste these examples to see the output:

```{r}
1:10
20:30
```

**Explanation**: The output shows that R has created a sequence of whole numbers between the start
and finish number.

To get a sequence with only even numbers, we can use the `seq` function, and set the `by` argument
to 2:

```{r}
seq(from=2, to=10, by=2)
```

You can set `by` to any number, including a decimal:

```{r}
seq(0, 27, by=3)
seq(0, 1, by=0.2)
```

If your sequence doesn't have a simple pattern, you can also write out the numbers by hand using the
`c()` function E.g.:

```{r}
c(1,40,92,188)
```

**Explanation**: `c(...)` is short for `combine`, so this command combines the numbers 1, 40, 92,
188 into a new sequence. This is sometimes called a **vector** in R-speak.

:::{.exercise}

Make some sequences which include:

-   Even numbers from 10 to 20
-   Numbers in the 8 times table less than 200
-   20 evenly spaced numbers between zero and 1 (including zero and 1)
-   The words "Wibble", "Wobble" and "Bobble"

`r hide("Show answers")`

We can use `seq` for numbers:

```{r}
seq(10,20,by=2)
seq(0,200, 8)
seq(0,1, by=1/19)
```

But we need to use c() for lists of words:

```{r}
c("Wibble", "Wobble", "Bobble")
```

`r unhide()`

:::

# Combinations of sequences {#expand-grid}

In designing experiments, or creating a grid of numbers for making predictions, we often want to create combinations of different categories which represent conditions or stimuli.

Imagine a hypothetical study with a test phase where participants are presented with multiple words,
in either red or green text, and shown at either the bottom or top of the computer screen.

The combinations look something like this:

```{r echo=F}
expand.grid(
   colour=c("Red", "Green"),
   position=c("Top", "Bottom"),
   word = c("Nobble", "Wobble", "Hobble")) %>%
   as_tibble() %>%
mutate(condition = row_number())%>% select(condition, everything())%>%
pander()
```

R provides quick ways of creating combinations of variables, using a command called `expand.grid`.

First, we need to create a sequence of each of the possible values for our categories:

```{r}
colours = c("Red", "Green")
positions = c("Top", "Bottom")
words = c("Nobble", "Wobble", "Hobble")
```

Then we can use `expand.grid` to give us all the possible combinations of these:

```{r}
expand.grid(colour=colours, position=positions, words = words)
```

**Explanation**: The `expand.grid` function has taken the items in the three input sequences
(colours, positions and words) and created a dataframe which contains all the possible combinations. We could save these to a file if we wanted to use them as part of our experiment.


:::  {.exercise}

Create some experimental designs of your own

1. Reproduce the experiment design above by copying and pasting

2. Adapt the commands to allow for an experiment where the word position could be either top,
   bottom, left or right. How many different conditions would there be in this case?

As a stretch task (this might be sligtly harder): How would you create a design where 
a sequence of 3 words is presented. Each word must be different, but each 3-word combination 
can be presented in red or green text, and at the top or bottom of the screen?


`r hide("Show solution")`

```{r}
expand.grid(colour=colours, position=positions, word1 = words, word2=words, word3=words) %>% 
  filter(
    word1 != word2 & word2 != word3 & word1 != word3
  ) 
```


**Explanation of the command**: There may be a neater way of doing this, but here we simply:

- Create all combinations of words, colors, spatial locations and positions in the sequence (1,2 or 3)
- Filter our rows where two of the words are the same



Another way to do the same thing using the `unique` and `length` functions might be:

```{r}
expand.grid(colour=colours, position=positions, word1 = words, word2=words, word3=words) %>% 
  rowwise() %>% 
  filter(length(unique(c(word1, word2, word3)))==3)
```
**Explanation**: Here we apply the sample filter, but using the `length` function to count the number of `unique` words among word 1, 2 and 3. The `rowwise` function is used to make R consider each row individually (it won't work without it because R tries to consider the whole of a column at once).



`r unhide()`
   
:::   
   


:::{.exercise}



Take the model from the main worksheet were we predicted grades from work hours for men and women.

In the main worksheet we created a dataframe by hand to tell `augment` what predictions we wanted.

Now try using `expand.grid` to make the new dataframe instead. 
For example, try making predictions for men and women who work 20, 25, 30, 35, or 40 hours per week. Make this dataframe using `expand.grid` and without using the `c()` function.
   
   
`r hide("Show how")`


```{r, messages=F, warning=F}
# setup and read data
library(tidyverse)
studyhabits <- read_csv('https://benwhalley.github.io/rmip/data/studyhabitsandgrades.csv')

# run the model
second.model <- lm(grade ~ work_hours * female, data = studyhabits)

# made the new data grid for predictions
prediction.grid <- expand.grid(work_hours=seq(20,40,5), female=c(TRUE, FALSE)) %>% 
  as_tibble
prediction.grid

# make the predictions
new.predictions <- broom::augment(second.model, newdata=prediction.grid)
new.predictions

# plot predictions from the model with SE
new.predictions %>% 
  ggplot(aes(work_hours, 
             y=.fitted, 
             ymin= .fitted-.se.fit, 
             ymax= .fitted+.se.fit, 
             color=female, group=female)) +
  geom_line() +
  geom_errorbar(width=.1) +
  geom_point()


# geom_ribbon is another way of showing the error of estimates
new.predictions %>% 
  ggplot(aes(work_hours, 
             y=.fitted, 
             ymin= .fitted-.se.fit, 
             ymax= .fitted+.se.fit, 
             color=female, group=female)) +
  geom_ribbon(alpha=.1, color=NA) +
  geom_line()
```


`r unhide()`

:::
   
   

# Clinical trial example


:::{.exercise}

Data from a clinical trial of Functional Imagery Training [@solbrig2019functional, FIT] are
available at <https://zenodo.org/record/1120364/files/blind_data.csv>. In this file, `group`
represents the treatment group (FIT=2; Motivational Interviewing=1). The `kg1` and `kg3` variables
represent the patients' weights in kilograms before and after treatment, respectively. Load these
data and complete the following tasks:

1. Plot the difference in weight between treatment groups at followup (`kg3`)

2. Create a plot to show whether men or women benefited most from the treatment (this will require
   some thinking about what goes on the y-axis, and perhaps some pre-processing of the data).

3. Create a plot to show whether older participants benefited more or less than younger ones (again
   this will require some thinking, and there are quite a number of different plot types which could
   be used, each with different pros and cons).
   
4. Run a linear model which is equivalent to the plot you created. Can you match the coefficients in the model output to the points and lines on your graph?

:::






# Preview: Multiple continuous predictors

In the previous example plotting our data was quite simple: We had an x and y axis, and used colour to show the third.

If we have two continuous predictors we _can_ do the same, using color or size to indicate the third dimension.
Here are two examples using some data on the income, murder and illiteracy rates of US states:



```{r}
# the data called state.x77 are in an old fashioned format, 
# so do this first to make them into a dataframe
states <- as_tibble(state.x77)
```


```{r fig.cap="3 dimensions using x, y and color"}
states %>% 
  ggplot(aes(x = Murder, y = Income, color=Illiteracy)) + geom_point()
```


```{r fig.cap="3 dimensions using x, y and point size"}
states %>% 
  ggplot(aes(x = Murder, y = Income, size=Illiteracy)) + geom_point()
```



Another alternative is to categorize one of the variables and make a faceted plot:

```{r}
states %>% 
  mutate(Illiteracy_categorised = cut(Illiteracy, breaks=3, 
                                      labels=c("Low illiteracy", "Medium illiteracy", "Highilliteracy"))) %>% 
  ggplot(aes(Murder, Income)) + 
   geom_point() +
  facet_wrap(~Illiteracy_categorised) +
  geom_smooth(method=lm, se=F)
```



**Explanation of the code** In the code above, the `cut` command is used to split the `Illiteracy` variable
into 3 categories (`breaks=3`) and give each category a label (`labels = c("Low", ...)`). We then use this categorical version to create a faceted plot for low, medium and high illiteracy values. 


----

If you look at these plots it's about possible to see that there is an **interaction** here: As `Illiteracy` increases, the relationship between `Murder` and `Income` changes (it starts positive but ends up slightly negative).


:::{.exercise}

Try describing the plot above in words.

Try re-plotting the data but with different values on different axes (e.g. swap x, y, color, size etc).

Which combination feels most 'natural' to you; that is, which is easiest to describe?


:::




## 3D plots

However, another way of displaying these data is to use a true 3D plot. An example is below:

```{r echo=F, include=T}
p1 <- plot_ly(data=states,
              x=~Murder, y=~Illiteracy, z=~Income, type = "scatter3d") 
p1
```


**Explanation of this plot**: Rather than a standard x/y plot, this one adds a third dimension, z. 
This means that to see all the data you need to click and drag on the image to rotate the axes.
It can take a bit of getting used to, but one tip is to make sure you keep track of which axis
is which as you move the viewing angle.


`r hide("Show me how to make the 3d plot")`

```{r, include=F, echo=T}
library(plotly)
plot_ly(data=states,
        x=~Murder, 
        y=~Illiteracy, 
        z=~Income, 
        type = "scatter3d") 
```

**Explanation of the code**: `ggplot` doesn't make 3D plots. Instead we use `plotly`, which is an
alternative graphing library. We use the `plot_ly` function, and pass it the `states` data
as the first argument. Then we define the x, y and z axes. Note that each variable name
from the dataset is prefixed by a `~` (tilde) character. This is required by `plot_ly`; it's
not consistent with `ggplot` which is a shame.

`r unhide()`




## Regression 'surfaces'

In the main session we saw we can represent the regression coefficients in a 2D plots. For example:

```{r}
states %>% 
  ggplot(aes(Income, Illiteracy)) + geom_point() + geom_smooth(method="lm", se=F)
```


This regression line looks problematic. The residuals (gaps from point to line) are quite large.

But, if we try to add `Murder` as another predictor of `Income` we can't represent the model as a simple line. Instead we need to think of the regression model as a 3d surface:


```{r, echo=F, include=T}
m <- lm(Income ~ Illiteracy *+ Murder, data=states)
axis_x <- with(states, seq(min(Murder), max(Murder), diff(range(Murder))/100))
axis_y <- with(states, seq(min(Illiteracy), max(Illiteracy), diff(range(Illiteracy))/100))
grid <- expand_grid(Murder=axis_x,Illiteracy=axis_y)
vals <- broom::augment(m, newdata = grid)

surf <- reshape2::acast(vals, Murder~Illiteracy, value.var = ".fitted")

p2 <- plot_ly(data=states,
              x=~Murder, y=~Illiteracy, z=~Income, type = "scatter3d") %>% 
  add_trace(z = surf,
                       x = axis_x,
                       y = axis_y,
                       type = "surface")
p2

```



:::{exercise}

Play with the interactive plot above. 

Is this 3d surface a good 'fit' to these data?

:::





In future sessions, as we increase the number of variables in our models, it will be difficult to 
keep hold of simple graphical equivalents to our models because it's hard to think in 5 or 6 or more dimensions.

It's good to remember this type of plot though. As we add more variables the `lm` function is still 
trying to minimise the gaps between the predicted line or surface and the actual data.




