## Cause and effect

```{r, echo=F, include=F}
library(tidyverse)
library(DiagrammeR)
```

### Confounding {#confounding-explanation}

The problem in the sample is that we missed out an important box: Smoking. This
is an example of an **_omitted variable bias_**, often called **confounding\_**.

Smoking causes cancer, and also makes it more likely that someone uses matches.
Because smokers use more matches than non-smokers, using more matches is
_associated with_ getting cancer, and we see this in the scatter plot. So,
although the scatterplot suggests matches cause cancer, we know it's wrong to
draw that conclusion.

When running observational studies (rather than experiments) the hardest job is
worrying about hidden variables that might influence our results in this way.


### Experiments and confounding {#explanation-experiments-confounding}

Good experiments make confounding impossible. To see how, consider this diagram
based on the example above:

```{r echo=F}
grViz("
    digraph {
        Matches -> Cancer
        Smoking -> Matches
        Smoking -> Cancer
        'Experimental Randomisation' -> Matches
    }
")
```

The digram represents an experiment in which we randomise our participants to
two experimental conditions, and followed them up over 2 years. At the end we
count how many participants in each group had cancer. The groups were:

-   Experimental: Participants are asked to strike 20 matches per day
-   Control: Participants are asked not to use matches

In the diagram we can see there are no arrows pointing at the box marked
`'Experimental Randomisation'`.

This means the Randomisation is **_uncaused_**. When we say something is random,
we mean there is no other variable which is causing it.


### Using other evidence to make assumptions {#explanation-assumptions-can-help}

It's important to use **observational** data wisely too. See @pearl2018book
chapter 5 for a great example of this: R.A. Fisher (the godfather of
experimental design) was wrong to dismiss the observational/correlational
evidence that smoking was harmful. Careful observation research linked smoking
with cancer and has saved millions of lives.
