---
title: 'Tending a Garden of Forking Paths'
author: 'Ben Whalley'
date: "March 2021"
bibliography: [references.bib]
biblio-style: apa6
link-citations: yes
output: 
  webex::html_clean:
    includes:
      after_body: footer.html
---



```{r, echo=F, include=F}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, cache=TRUE, comment=">", message=FALSE)
library(tidyverse)
library(webex)
library(pander)
theme_set(theme_minimal())
```



In the seminar we discussed Simmons et al. (2011) and  Gelman and Loken (2013). The data from Simmons et al are available at: <https://zenodo.org/record/7664#.YFSoRi-l2CM>

A short cut to download and unzip the data from the repository is shown below:

```{r}
# download the file directly from R
dataurl <- "https://zenodo.org/record/7664/files/FalsePositive_Data_in_.txt_format.zip?download=1"
download.file(dataurl, "simmons.zip")

# unzip the file with R
# this places the data files from the zip into the 
# working directory (so use an RStudio project)
unzip('simmons.zip')
```


This creates 2 data files and a codebook in the current directory (e.g. in your RStudio server account).

You can read in the data like so:

```{r}
# note these are tab separated, not comma, so use read_tsv
study1 <- read_tsv('Study 1 .txt')
study2 <- read_tsv('Study 2.txt')
```

And take a peek at them:

```{r}
study1 %>% glimpse
```


The codebook for these data is:

```
Variable	Description
age_days	Days since participant was born (based on their selfreported birthday)
dad	What's is/would-be your father's age today (in years)
mom	What is/would-be your mother's age today (in years)
female	Subject is a woman?
root	"Did they get correclty what the square root of 100 is? (1-yes, 0-no)"
bird	Imagine a restaurant you really like offered a 30% discount for dining between 4 pm and 6 pm.  How likely would you be to take advantage of that offer? ? (1) Very unlikely ?(7) Very likely
political	"In the political spectrum, where would you place yourself (1-Very liberal, 2-liberal, 3-centrist, 4-conservative, 5-very conservative)"
quarterback	"if you had to guess who was chosen the quarterback of the year in Canada last year, which of the following four options would you choose? 
1 Dalton Bell; 2 Daryll Clark; 3 Jarious Jackson ;4 Frank Wilczynski"
olddays	"How often have you referred to some past part of your life as ""the good old days""?
11 Never;12 almost never;13 sometimes;14 often;15 very often"
potato	WHICH SONG were they assigned to?
when64	
kalimba	
feelold	"How old do they feel (1 very young, 2 young, 3 neither young nor old, 4 old, 5 very old)"
computer	Computers are complicated machines (1-strongly disagree?5-strongly agree)
diner	"Imagine you were going to a diner for dinner tonight, how much do you think you would like the food?  (1 Dislike extremely?9 Like extremely)"
cond	Condition
aged365	age in years

```




# Grow a garden of forking paths!



```{r, echo=F, include=F}
study2.subset <- study2 %>% filter(when64==1|kaimba==1)
m1 <- lm(aged365 ~ cond + dad, data=study2.subset)
summary(m1)

m1.bf0 <- BayesFactor::lmBF(aged365 ~  dad, data=study2.subset)
m1.bf <- BayesFactor::lmBF(aged365 ~ cond + dad, data=study2.subset)
m1.bf/m1.bf0
```


:::{.exercise}


- Can you reproduce the result reported from study 1?
- What's the smallest *p* value you can find that supports a publishable story?
- How easy is it to 'disrupt' the finding reported in the paper?

:::




