
XXX THIS SHOWS UPLOADING NOT LOADING FROM WEB




### Random samples

In our example, each of the experimental conditions is allocated between-participants, which (as
you'll learn elsewhere in the programme) means we need recruit quite a large number of participants.
In this case we can imagine we might need at least 250 participants.

Allocating participants to condition is another boring task R can help with. The trick is to combine
sequences with randomness and random sampling.

For example, we might create a sequence of the numbers from 1:10:

```{r}
1:10
```

The `sample` function lets us take a random sample from this sequence:

```{r}
sample(1:10)
```

**Explanation** `sample` has shuffled the input sequence and gives us the original numbers (1...12)
in a random order.

---

If we want to take a larger sample, longer than the input sequence, we have to add the text
`replace=TRUE`:

```{r}
sample(1:10, size = 20, replace=TRUE)
```

**Explanation**: Using `replace=TRUE` is like picking a number from a hat, but then putting the
chosen number back into the hat so it can be picked again. This allows us to make longer random
sequences where numbers in the initial sequence are repeated.

### In-use: Randomising participants to conditions/groups

If we combine our experimental design from above with this idea of randomness we can randomise
participants to conditions in our experiment.

If we look again at our design we can see there are 12 possible combinations:

```{r}
expand.grid(colour=colours, position=positions, words = words)
```

We can use `sample` to make sure we allocate participants randomly to conditions, as they are
recruited:

```{r}
sample(1:12, size=250, replace=TRUE)
```

So, we would allocate participants to their condition in this order as they are recruited.

Alternatively, we could simply run this every time a new participant showed up:

```{r}
(new_allocation <- sample(1:12, size=1))
```

#### Task: Randomise your own participants {.exercise}

1. Re-use the design you created above, and randomise 5 participants to conditions within it.

2. Try and create a design which might work for your own study (either this year, based on your UG
   project, or on any other study you might like to run).

As an optional stretch task:

1. [Read this extension, which shows how to sample from datasets as well as sequences](#sample_n)


# Extension tasks {#optional-extensions}

### Missing data: Try the same code on another dataset {.exercise}

The `mice` package includes example datasets with missing data. We can look at one of these like so:

```{r}
mice::boys %>% glimpse
```

1. Identify the patterns of missing dataset in the `mice::boys` dataset.

2. How many participants provided complete data?



<!-- mice::boys %>% mice::md.pattern() -->
