---
title: "Notes/code in response to queries in QA session, 19th April"
output: html_notebook
---




```{r}
library(tidyverse)
```



# Session 1


## Formatting numbers with `paste`
There were a number of queries about formatting numbers and combining them in a single column.

These were the examples I gave:

```{r}
mtcars %>% 
  # select just what we want
  select(mpg, cyl) %>% 
  # use paste0 to  combine into text
  # note paste0 is like paste but doesn't include spaces
  mutate(finaloutputcolumn = paste0(mpg, " (", cyl, ")"))

```


## Using `str_interp`

The `str_interp` function is a more advanced version of `paste`.  This is NOT specifically required for the assessment, but in response to questions about it I gave this example code:

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(mean_wt=mean(wt)) %>% 
  # remember to use rowwise or you will end up with all the data pasted into every row
  rowwise() %>% 
  # the $[.2f]{wt} part is explained below
  mutate(mean_wt_fixed_width = str_interp("Mean =  $[.2f]{mean_wt}"))

```


In the code above, I write: `$[.2f]{wt}`.  Broken down this means:

- `$` -- we are going to incude a column variable
- `[.2f]` -- we want to format the number to 2 dp, and be fixed width (so include trailing zeros). This part is optional
- `{wt}` -- this gives the name of the column variable




In this example I showed how we could break up code into 2 steps to simplify:

```{r}
mtcars_summ <- mtcars %>% 
  pivot_longer(everything(), names_to="Variable") %>% 
  group_by(Variable) %>% 
  summarise(Mean=mean(value), SD=sd(value)) 

mtcars_summ %>% 
  rowwise() %>% 
  mutate(
    `Mean (SD)` =  str_interp("$[.2f]{Mean} ($[.2f]{SD})" )) %>% 
  select(-Mean, -SD)
```



## Removing repetition in tables

Max asked about avoiding repetition in columns. I got stuck trying to make it work when live, but the working code is below:
This is NOT required for the assessment. 


```{r}
mtcars %>% 
  group_by(am, cyl) %>% 
  summarise(wt=mean(wt)) %>% 
  arrange(am, cyl) %>% 
  group_by(am) %>% 
  mutate(am  = ifelse(row_number()==1, am, ""))
```



## Sorting text and factor variables

Hannah asked about sorting. Sometimes alphanumeric sorting isn't what we want for categorical variables. We
can fix this by using an explicit order when creating a factor variable:

```{r}
# sorting alphabetically
tibble(name = c("Apple", "Orange", "Banana")) %>% 
  arrange(name)


times_in_order <- c("Time 1", "Time 2", "Pre test",  "Afterward")

example_data <- tibble(time = sample(times_in_order, 12, replace=T))

# data is in random order
example_data

# now alphabetical sorting is not what we want
example_data %>% 
  arrange(time)


# we convert time to a factor and explicitly tell R the ordering of the `levels` in the factor
example_data %>% 
  mutate(time = factor(time, levels = times_in_order)) %>% 
  # now when we sort we get the order we want
  arrange(time)
  
```



You can also see this link for more info: http://www.cookbook-r.com/Manipulating_data/Changing_the_order_of_levels_of_a_factor/





# Splitting of tables


This is useful to make sure wide tables don't get split


```{r}
library(pander)

# do this at the top, e.g. when you load tidyverse
panderOptions('table.split.table', 100)

```



Or do it per-table:




```{r}



mtcars %>% head(3) %>% pander(split.tables=100)

mtcars %>% head(3) %>% pander()
```







# Session 2



Converting factors to numeric:

```{r}
sleep %>% 
  mutate(grp_numeric = as.numeric(group))
```

Remember this won't always make sense — if the variable is truly categorical using it as a linear predictor in a regression would be inappropriate.



## An example of pivoting to compare variables

```{r}

tibble(person = c(1,1,2,2), time = c("Time1", "Time2", "Time1", "Time2"), 
       score = c(11,12,32,33)) %>% 
  pivot_wider(names_from = time, values_from=score) %>% 
  mutate(d = Time1 - Time2) %>% 
  summarise(mean(d))

```










