---
title: 'Data wrangling 2'
author: 'Ben Whalley'
date: "November 2021"
bibliography: [references.bib]
biblio-style: apa6
link-citations: yes
output:
  webex::html_clean
---


```{r, include=F}
source('_first_chunk.R')
```






# Before you start

:::{.exercise}

Continue to work in the `rmd` file you made in the last session, in your datafluency directory (it should be called `data-wrangling.rmd` or similar).


:::





# Overview

In this session we cover 3 common tasks when tidying up real datasets:

1. Recoding data (e.g. questionnaire responses) from text to numeric values
2. Separating 'untidy' variables into tidy, long-form data:

Additionally, as an optional extension activity, we cover:

3. Joining two sources of data (e.g. two spreadsheets) into a single dataframe





# Separating variables

Sometimes we need to separate 'untidy' variables into tidy, long-form data.


<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/NRaKlYGaXEs" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


The code below generates simulated data for 100 individuals at three time points. The format is
similar to the way you might record experimental data in a spreadsheet.

```{r}
N <- 100
repeatmeasuresdata <- tibble(person = 1:N,
                              time_1 = rnorm(N),
                              time_2 = rnorm(N, 1),
                              time_3 = rnorm(N, 3))

repeatmeasuresdata %>% head(8)
```

This variable, `repeatmeasuresdata`, is in **wide** format. Each row contains data for one participant, and each
participant has three observations.

As [we saw previously](data-wrangling1.html#pivotlonger), we can *pivot* --- i.e., reshape --- the data into longer format like so:

```{r}
repeatmeasuresdata %>%
  pivot_longer(starts_with("time")) %>%
  arrange(person, name) %>%
  head(8)
```

The problem we have now is that `name` contains text which describes at which time the
observation was made. We probably want to store a *number* for each time-point, so we can make a plot with time
on the x axis.

The `separate` command separates a single character column (`name`) into multiple columns.
Rather than have a column with labels of the form 'time_1', it can create two columns, with labels
'time' and '1' in each.

```{r}
# convert to long form; extract the `time` as a new numeric column
longrepeatmeasuresdata <- repeatmeasuresdata %>%
  pivot_longer(starts_with("time")) %>%
  separate(name, into = c("variable", "time"))

longrepeatmeasuresdata %>% head
```


Now the data are in long format, we can plot the points over time:

```{r}
longrepeatmeasuresdata %>%
  sample_n(30) %>%
  ggplot(aes(x=time, y=value)) +
  geom_point()
```


### How does R know where to split the text?


In the example above, `separate` split data like `"time_1"`, `"time_2"` etc into two columns: `variable` and `time`.

Q: How did it know to use the underscore (`_`) to split the data?

A: The default is to split on anything which is not a letter or number. So `_` or a space, or `,` would all work.


Sometimes though we need to tell R explicitly what to use to sepatate the values.

If we had a column of email addresses we could split `ben.whalley@plymouth.ac.uk` into the username (e.g. `ben.whalley`) and domain name (`plymouth.ac.uk`) using the `@` symbol.

To do this we just write `sep="@"` when we use separate.



:::{.exercise}

The `messy_exp` dataset in `psydata` contains simulated RT data on 100 participants in 2 conditions (A and B) at three time points (1, 2, and 3).

- Use the `separate()` function to split up the `condition` variable in this dataset and draw the following plot:

```{r, echo=F}
messy_exp %>% 
  separate(condition, into=c("participant", "condition", "time")) %>% 
  ggplot(aes(time, rt, color=condition)) + 
  geom_boxplot(width=.5) + 
  labs(x="Time", y="Reaction time (ms)", color="Condition")
```
`r hide("Show the code")`

```{r, echo=T, eval=F}
messy_exp %>% 
  separate(condition, into=c("participant", "condition", "time")) %>% 
  ggplot(aes(time, rt, color=condition)) + 
  geom_boxplot(width=.5) + 
  labs(x="Time", y="Reaction time (ms)", color="Condition")
```

`r unhide()`

:::


:::{.exercise}


1. This file contains sample contact and address data for 100 people: <https://letterhub.com/wp-content/uploads/2018/03/100-contacts.csv>


  - Read the data into R (you can either use the URL above directly inside the `read_csv()` function, or download then re-upload the data to the server to do this)

  - Use the `separate` function to make a new variable which contains the *domain name* of these contacts' email address (e.g. yahoo.com, hotmail.com)

> Note, you will need to use `sep="@"` to split the email addresses at the `@` symbol

2. Use the `distinct` and/or `count` functions on the new variable you create containing the domain name. Look them up in the help file if you don't know which to use to answer these questions:

  - How many people had a Gmail account?
  - Which domains had more than 10 users?


`r hide("Show workings")`

```{r}
# read the data directly from the URL
contacts <- read_csv('https://letterhub.com/wp-content/uploads/2018/03/100-contacts.csv') %>% 
  separate(email, into=c("user", "domain"), sep ="@")  # uses the @ symbol as a separator
```

```{r}
# how many _different_ domains are there?
contacts %>% 
  distinct(domain) %>% 
  count()
```

```{r}
# how many people use gmail
contacts %>% 
  count(domain) %>% 
  filter(domain=="gmail.com")
```

```{r}
# which domains had more than 10 users?
contacts %>% 
  count(domain) %>% 
  filter(n > 10) 
```

`r unhide()`


:::






# Questionnaire data

## Importing from Excel

You [should always keep your data in csv format](#why-use-csv). Don't use Excel
or other proprietary formats to store data unless absolutely necessary.

Unfortunately, some tools (including Office 365 Forms) provide data in `xlsx` format, and
for this we need to use a special function in R. You can import this data to RStudio by:

-   Uploading the `xlsx` file to the RStudio server, as you have done before
-   Loading the `rio` package
-   Using the `import()` function in that package, instead of `read_csv()`

The example data linked below were exported from an Office 365 Form, so they are in a similar format 
to that which you might expect using other online survey tools.


- **`r xfun::embed_file('data/sweets.xlsx', text="Download 'sweets' example in xslx format")` (`sweets.xlsx`)**


```{r, echo=T, eval=F, include=F}
# import an excel file using rio
library(rio)
sweets <- import('sweets.xlsx')
```

```{r, echo=F, include=F}
# chunk above is 'faked' to hide data/ path which confused students in the first year we ran this
library(rio)
sweets <- import('data/sweets.xlsx')
```

We can look at the first few rows of the data, using the `head` command:

```{r}
sweets %>% head()
```


<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/oKpdIxG-7ak" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>




:::{.exercise}


- Import the `r xfun::embed_file('data/sweets.xlsx', text="sweets data")` into RStudio, as shown above

- Save it to a new variable called `sweets`

:::




## Tidying questionnaires

When we look at the imported data it's useful to note:

1. There are extra columns we don't need (at least for now).

2. Some of our variable names are very long and annoying to type (for example
   `How much do you like sweets?` is the name of one of our columns).

3. Our responses are in **text** format, rather than as numbers. For example,
   the data say `"I don't like them"` or `"I'm neutral"` rather than numbers
   from a 1-5 scale.

4. We may have other data on the same participants that are stored in separate files.


We need to sort each of these problems to make things more manageable for our
analysis.




## Selecting and renaming 

> Remember, R makes using columns with spaces or other special characters very hard. We want to avoid this.

### Selecting 

To use columns with spaces in we must 'escape' the spaces and **let R know they are part of the name** rather than a gap between two different names.

This video shows how (or read below):

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/aIMgsj5hTVA" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


To escape spaces and use columns with long names we use the backtick character (the
backwards facing apostrophe) to *wrap* the column name.  

**In general, if your columns contain spaces or other odd characters like hyphens or question marks then you will need to wrap them in backticks.**




### Renaming  {#renaming}

Some of the imported variable names in the `sweets` data are long and awkward to use.

Most researchers would rename these variables, to make them more usable in R code.

You can rename variables like this:

```{r, eval=F}
datasetname %>% 
  rename(NEW_COLUMN_NAME = OLD_COLUMN_NAME)
```

So for this example:

```{r}
sweets %>%
  rename(
    like_sweets = `How much do you like sweets?`,
    like_chocolate = `How much do you like chocolate`,
  )
```

**Explanation of the code**: We used `rename` to change the names of our
variables. We needed to wrap the long names of the questions in 'backtick'
symbols to make sure R understood it was a single column name.

You should create a new variable to save the renamed dataset (with a descriptive name for use later on):

```{r}
# create a new variable containing the renamed dataset
sweets.renamed <- sweets %>%
  rename(
    like_sweets = `How much do you like sweets?`,
    like_chocolate = `How much do you like chocolate`,
  )
```



:::{.exercise}

1. Create a copy of the sweets data in which you have selected only the two columns with long names.

1. Create a second copy of the data where you have renamed the columns with long names to something short, and without spaces.

:::



## Recoding text {#using-recode}

We noticed above that our responses were stored as text labels like
`"I don't like them"` rather than on a numeric scale. This makes it hard to use in
an analysis.

**We need to _recode_ the text variables into numeric versions.**

### How to do it

First we must tell R what number we want to use for each text label. That is, we
create a **_mapping of numbers to labels_**.

This takes a few steps:

1. Check **exactly** what the text values are which need to be mapped. 
2. Make a **mapping variable** which assigns each text value a number value
3. Use the `recode` function with `mutate` to create a new, numeric column


This video walks you through the steps below:

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/vaGrKPIHN4Q" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>


### Step 1: Check EXACTLY what text labels we have

To check which labels we need to recode, I select the column in question and use the `unique()` function.

```{r}
# check exactly what text values are in the dataset?
sweets %>% 
  select(`How much do you like sweets?`) %>% 
  unique()
```

:::{.exercise}

Do the same to find out the possible values of the `How much do you like chocolate` column.

`r hide("Show answer")`

```{r}
sweets %>% 
  select(`How much do you like chocolate`) %>% 
  distinct()
```

`r unhide()`

:::



### Step 2: Make a mapping variable

We do this by creating what R calles a **named vector**, which is a special kind of list.

To make a named vector we use the the `c()` function. The letter `c`
here just stands for 'combine' --- i.e. 'combine these things into a list'.

This is a simple example:


```{r}
mapping.list <- c("No" = 0, "Yes" = 1)
```


We could then use this mapping to recode a column of data which contained the words "No" or "Yes


<br>


*A useful trick* when creating your own mappings is to use R to do the formatting for you (see the video above for a demo).

Re-using the code from the previous step, we use `unique()` to show us the *unique values* for the quuestion about sweets.

We then pipe the result to the `paste()` and `cat()` functions, like this:

```{r}
# the hack we use as a short-cut to creating a mapping variable
sweets %>% 
  select(`How much do you like sweets?`) %>% 
  unique() %>% 
  paste() %>% cat()
```

**Explanation of the output**: Using `paste` and `cat` is a bit of a hack. When we run this code we see the output `c("I don't like them", "I'm neutral", "I like them")`. This is a list of the values in the `sweets` data for this question, formatted in a way that will be useful to us in the next step.


We then copy and paste this output into a NEW code block, and edit it to assign our mappings:

```{r}
preference.mappings <- c("I don't like them" = -1, "I'm neutral" = 0, "I like them" = 1)
```

**Explanation of the code**: We used the previous output to create a mapping. By adding the `= -1` and `= 0` etc, we have told R what value we want to assign for each label.



:::{.tip}

Q: How do you know what number values to assign?

A: It doesn't matter, provided:

- The intervals between each options are the same and 
- Each text value has a different number

So, if we had a Likert-scale ranging from "Completely agree" to "Completely disagree" in 7 increments, we could score this from `0 - 6` or `1 - 7`, or `-3 - 3`. These would all be fine.


:::



### Step 3: Use the mapping variable to recode the column

We can use our new mapping with the `mutate` and `recode` functions to make a **new column**, containing numbers rather than text:

```{r}
sweets.recoded <-  sweets %>%
    # use recode to convert text response using preference.mappings
    mutate(
        like_sweets_numeric =
            recode(`How much do you like sweets?`, !!!preference.mappings)
    )
```

We can see this new column if we use `glimpse`:

```{r}
sweets.recoded %>% glimpse()
```


**Explanation of the code**: 
- The start of the first line is `sweets.recoded <-` which means
make a new variable called `sweets.recoded`.
- Then we use `mutate` to create a **new** column called `like_sweets_numeric`. 
- We make this column using `recode` on the question about liking sweets. 
- We use the `preference.mappings` mapping to specify what numeric score to give each of the text values.

**Watch out for the exclamations marks`!!!`**: In the code there are
[three exclamation marks](#explain-exclamationmarks), `!!!`, before the mapping; make sure you do the same.



## Summary/video explanation

This is one of the trickiest bits of R code we use. I've included an annotated video of just these steps as a reference below:

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/DCeAlHUZsF0" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>





:::{.exercise}

- Use this three-step process to create a recoded version of the `like_chocolate` variable.

- Remember to watch the video at the start of this section, or the short version in the green box above, if anything is unclear.

:::





## Reverse codings

Watch out for "reverse coded" items. Imagine a questionnaire that includes two items:

- "I really love sweets"
- "I really hate sweets"

Participants respond on a scale of "Agree", "Disagree", and "Neutral".


Using the same mapping to both these questions would not be appropriate. Instead, we should use
two separate mappings for the different questions to ensure the ***direction*** of the 
coding has the same meaning:


```{r}
# mapping for items like "I really love sweets"
agree.disagree.responses <- c(
            "Agree" = 3,
            "Neutral" = 2,
            "Disagree" = 1)
```



```{r}
# reversed mapping for items like "I really hate sweets"
agree.disagree.responses.REVERSED <- c(
            "Agree" = 1,
            "Neutral" = 2,
            "Disagree" = 3)
```


Coded in this way, scores for the questions _could_ be added or averaged to provide a summary score.



## Multiple new columns

You can create multiple new columns at once using mutate, to recode each of our
variables:

```{r}
sweets.recoded  <- sweets.renamed %>%
    # create two columns at once using mutate
    mutate(
        like_sweets_numeric = recode(like_sweets, !!!preference.mappings),
        like_chocolate_numeric = recode(like_chocolate, !!!preference.mappings)
    )
```

**Explanation of the code**: We combined the two examples above to create 3 new
columns with recoded (numeric) values. We saved this to the variable called
`sweets.recoded`.

  




# Combining scores {#summary-score-of-questionnaire}

Often questionnaires are designed to make repeated measurements of the same phenomena, which can then be summed or averaged to 
create a more reliable measure.

We've already seen how  `mutate()` creates a new column. We can use this again to create the *sum* of
responses to both questions:

```{r}
sweets.recoded %>%
  # mutate to create column containing sum of both recoded questions
  mutate(liking = like_sweets_numeric + like_chocolate_numeric) %>%
  select(starts_with('like'))
```


**Explanation of the code**: 
- We added both the questions which asked about 'liking' together. This
created a new column containing the combined score, called `liking`. 
- In the third line we selected only columns with names that *started with* the word 'like', to make the output easier to view.




# Consolidation activity

:::{.exercise}

Use this example dataset: `r xfun::embed_file('data/sleep.xlsx', text="sleep.xlsx")`

- Read in the data using the `import` function in the `rio` package
- Rename the long column names to something short, and without spaces
- Recode at least three of the columns with data about sleep quality to be numeric
- Save the result of this work in a new variable called `sleep.tidy`
- Pivot the recoded variables and make a boxplot of responses to these questions
- Create a summary score called `sleep_quality` which is the sum of these recoded questions (use mutate)
- Create a [density plot](#makingdensity) of this summary score and interpret what you see (describe the pattern in plain English)


`r hide("Show complete code example")`

```{r, include=F, echo=F}
sleep <- import('data/sleep.xlsx')
```


```{r, eval=F}
# this block assumes you have already uploaded the data to the server
library(rio)
sleep <- import('sleep.xlsx')
```

```{r}
# used to check what response values are in each question
sleep %>% 
  select(`My sleep is affected by my study commitments`) %>% 
  unique() %>% paste %>% cat

sleep %>% 
  select(`My electronic device usage negatively affects my sleep`) %>% 
  unique() %>% paste %>% cat
```


```{r}
# we will use the same mapping for both questions because they have the same responses
sleep.map <- c("Agree"=2, 
               "Somewhat agree"=1, 
               "Somewhat disagree"=-1, 
               "Disagree"=-2, 
               "Neither agree nor disagree"=0, 
               "Strongly agree"=3, 
               "Strongly disagree"=-3)


sleep.tidy <- sleep %>% 
  # now we recode the two text variables (we only need use mutate once though)
  mutate(
    sleep_study = recode(`My sleep is affected by my study commitments`, !!!sleep.map), 
    sleep_electronic = recode(`My electronic device usage negatively affects my sleep`, !!!sleep.map)
  )

```

```{r}
# now we can pivot longer to make a plot
sleep.tidy %>% 
  pivot_longer(c(sleep_study, sleep_electronic)) %>% 
  ggplot(aes(name, value)) + geom_boxplot()
```

And make a summary score combining both questions

```{r}
sleep.tidy.withsumary <- sleep.tidy %>% 
  # and create the summary score
  mutate(sleep_quality = sleep_study + sleep_electronic ) 
  
# check the result. it looks ok
sleep.tidy.withsumary %>% glimpse
```


```{r}
# finally, make the requested density plot
# make the density plot
sleep.tidy.withsumary %>% 
  ggplot(aes(sleep_quality)) + 
  geom_density()
```



`r unhide()`


:::







# Joining data

:::{.tip}

This is an optional extension section. It's not needed for the assessment.

:::



Often we will have data from two different sources which must be combined. This section explains how to do that.


<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/D32xVe3-YVg" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
          
          

```{r, echo=F, include=F}
personal_data <- tibble(Identifier=c("123ABC", "456DEF", "678FGH"),
  Name=c("Ben", "Helen", "Esther"))
research_data <- tibble(Identifier=c("123ABC", "456DEF", "678FGH"),
  `Worst ever record purchase`=c("Europe, the final countdown", "Spice Girls Wannabe", "One Direction - Night changes"))
```


Imagine we have two datasets like this, one containing identifying information:

```{r, echo=F}
personal_data %>% pander::pander()
```

And a second dataset containing sensitive questionnaire data:

```{r, echo=F}
research_data %>% pander::pander()
```
> Note that the `Identifier` column is common to both datasets.

We can use a new `tidyverse` function which joins these two datasets into a combined table:

```{r}
left_join(personal_data, research_data, by="Identifier")
```

#### Explanation of `left_join`

The `left_join` function takes two dataframes as input. We call the
first input the 'left hand side', and the second input the 'right hand side'.

Both left and right sides have to **share at least one column**. In this case it's
called `Identifier`.

When the left and right hand side match then `left_join` copies the extra columns *from the right hand side into the left hand side*. 

**It returns a combined dataframe.**



Any extra rows in the right hand side which don't match the left hand
side are dropped. This might happen if we have research data for people we don't have personal data for.
There are [other functions like `full_join`](https://dplyr.tidyverse.org/reference/join.html#join-types)
which do this a bit differently, but we don't need them for now.


:::{#joining-data-task .exercise}

- Have a look at the `heroes_meta` and `heroes_personal` datasets, in `psydata`.

- Adapt the code for `left_join()` shown above to combine these datasets.

- Pipe the result of `left_join()` to `count()` to check how many rows are in the final dataset

- Does it matter which order you join the datasets in? Can you explain the number of rows in the resulting dataset?

- Save the combined dataset into a new csv file using `write_csv()`. Check you can find this in the RStudio files pane, and download a copy to be sure.


`r hide("Show answers")`

Do the join:

```{r}
left_join(heroes_meta, heroes_personal)
```


Count the rows:

```{r}
left_join(heroes_meta, heroes_personal) %>% count()
left_join(heroes_personal, heroes_meta) %>% count()
```

We get slightly different numbers depending on which order we join the data in.

This is because `left_join` only keeps rows in the right hand dataset IF they have a match in the left hand side.
If the number of rows in the left and right hand sides differ, you can get a different result.

Other functions like `inner_join` and `full_join` take different approaches. Type `?dplyr::inner_join` for an explanation.

This code would export a csv file to the same location as your `rmd` file:

```{r, eval=F}
left_join(heroes_meta, heroes_personal) %>% rio::export('combined.csv')
```


`r unhide()`


:::



# Extra bits


## Making a density plot {#makingdensity}

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/0r8rRvNF73c" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>






